---
title: "Class 8 Mini Project "
author: "Angela Abraham"
date: "2025-10-23"
output: pdf_document
---

#Save the input data file into the Project directory
```{r}
fna.data<-"WisconsinCancer.csv"
```

#Input the data and store as wisc.df
```{r}
wisc.df<-read.csv(fna.data, row.names=1)
wisc.df
```


```{r}
wisc.df$diagnosis
```

#Use -1 to remove the first column
```{r}
wisc.data<-wisc.df[,-1]
```

#Create the diagnosis vector to use later
```{r}
diagnosis<- as.factor(wisc.df$diagnosis)
colnames(wisc.df)
```

>Q1. How many observations are in this data?

```{r}
nrow(wisc.df)
```


>Q2. How many of the observations have a malignant diagnosis?

```{r}
table(diagnosis)
sum(diagnosis=="M")
```

>Q3. How many variables/ features in the data are suffixed with _mean?

```{r}
sum(grepl("mean$", colnames(wisc.df)))
```

#Principal Component Analysis

## Performing PCA

#Check column means and standard deviations 
```{r}
colMeans(wisc.data)
apply(wisc.data, 2, sd)
```

#Perform PCA on wisc.data 
```{r}
wisc.pr<-prcomp(wisc.data, scale.=TRUE)
```

#See the summary results
```{r}
summary(wisc.pr)
```

>Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

```{r}
summary(wisc.pr)
summary(wisc.pr)$importance[2,1]
```

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

```{r}
which(summary(wisc.pr)$importance[3,]>0.70)[1]
```

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

```{r}
which(summary(wisc.pr)$importance[3,]>=0.90)[1]
```

```{r}
plot(summary(wisc.pr)$importance[3,], type="b", xlab="Principal Component", ylab="Cumulative Proportion of Variance", main="Variance Explained by Principal Components")
```

##Interpreting PCA results

>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```

This plot is poorly designed. It is clustered and there is too much names/ numbers all over the plot. This is difficult to understand since it's clustered. 

#Scatter plot observations by components 1 and 2
```{r}
plot(wisc.pr$x[,1:2], col=ifelse(diagnosis=="M","violetred3","black"), xlab="PC1", ylab="PC2")
```

>Q8. Generate a similiar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(wisc.pr$x[,c(1,3)], col=ifelse(diagnosis=="M", "violetred3","black"), xlab="PC1", ylab="PC3")
```

# We can see a clear separation between maligannt (violet) and benign (black) samples. This is a cleaner plot that separates the two subgroups.

#Create a data.frame for ggplot
```{r}
df<-as.data.frame(wisc.pr$x)
df$diagnosis<-diagnosis
```

#Load the ggplot package
```{r}
library(ggplot2)
```

#Make the scatterplot colored by diagnosis
```{r}
ggplot(df)+aes(PC1, PC2, col=diagnosis)+geom_point()
```

#Calculate the variance of each component
```{r}
pr.var<-wisc.pr$sdev^2
head(pr.var)
```

#Variance explained by each principal component: pve
```{r}
pve<-pr.var / sum(pr.var)
```

#Plot variance explained for each principal component
```{r}
plot(pve, xlab="Principal Component", ylab="Proportion of Variance Explained", ylim=c(0,1), type="o")
```

#Alternative scree plot of the same data, note data driven y-axis
```{r}
barplot(pve, ylab="Present of Variance Explained", names.arg=paste0("PC", 1:length(pve)), las=2, axes=FALSE)
axis(2, at=pve, labels=round(pve,2)*100)
```

##ggplot based graph
#install.packages ("factoextra")
```{r}
options(repos=c(CRAN="https://cloud.r-project.org"))
install.packages ("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels=TRUE)
```

>Q9. For the first principal component, what is the component of the loading vector (i.e wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

>Q10. What is the minimum number of principal components required to explian 80% of the variance of the data?

```{r}
which(summary(wisc.pr)$importance[3,]>=0.80)[1]
```

## Hierarchical clustering

#Scale the wisc.data data using the "scale()" function
```{r}
data.scaled<-scale(wisc.data)
```

```{r}
data.dist<-dist(data.scaled)
```

```{r}
wisc.hclust<-hclust(data.dist, method="complete")
```

## Results of hierarchical clustering

>Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
rect.hclust(wisc.hclust, k=4, border="red")
abline(h=30, col="red", lty=2)
```

## Selecting number of clusters

```{r}
wisc.hclust.clusters<-cutree(wisc.hclust, k=4)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```

>Q12. Can you find a better cluster vs. diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
for (k in 2:10) {
  cat("\nNumber of clusters:", k, "\n")
  print(table(cutree(wisc.hclust, k = k), diagnosis))
}
```

## Using different methods

>Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
par(mfrow=c(2,2))
wisc.hclust.single   <- hclust(data.dist, method = "single")
wisc.hclust.complete <- hclust(data.dist, method = "complete")
wisc.hclust.average  <- hclust(data.dist, method = "average")
wisc.hclust.ward     <- hclust(data.dist, method = "ward.D2")
```

```{r}
plot(wisc.hclust.single,   main = "Single Linkage")
plot(wisc.hclust.complete, main = "Complete Linkage")
plot(wisc.hclust.average,  main = "Average Linkage")
plot(wisc.hclust.ward,     main = "Ward.D2 Linkage")
```

#Ward D2 method has the clearest separation between the malignant (M) and benign (B) diagnoses. The graph was more structured in clusters compared to the others.

##Combining methods

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2")
grps<- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
table(grps, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

```{r}
g <-as.factor(grps)
levels(g)
```

```{r}
g<-relevel(g,2)
levels(g)
```

#Plot using the re-ordered factor
```{r}
plot(wisc.pr$x[,1:2], col=g)
```

>Q15. How well does the newly created model with four clusters separate out the two diagnoses?

```{r}
wisc.pr.hclust.clusters<-cutree(wisc.pr.hclust, k=2)
table(wisc.pr.hclust.clusters, diagnosis)
```
# Separates well

>Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
wisc.km <- kmeans(scale(wisc.data), centers=2, nstart=20)
table(wisc.km$cluster, diagnosis)
table(wisc.hclust.clusters, diagnosis)
```

#Sensitvity / Specificity

>Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

```{r}
sensitivity<-188/(188+24)
print(paste("Sensitivity:", round(sensitivity, 3)))
```

```{r}
specificity <-329 / (329+28)
print(paste("Specificity:", round(specificity,3)))
```
# PCA + hierarchical clustering model has the best sensitivity & specificity

#Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

>Q18. Which of these new patients should we prioritize for follow up based on your results?

#Patient 2


