---
title: "Halloween Mini Project"
author: "Angela Abraham"
date: "2025-10-30"
output:
  pdf_document: default
---

## Importing candy data

```{r}
candy_file<-"https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy<-read.csv(candy_file, row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
#85 candy types

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

#38 candy types in the dataset

## What is your favorite candy?

```{r}
candy["Twix",]$winpercent
```

>Q3. What is your favorite candy in the dataset and what is it's winpercent value?

```{r}
rownames(candy)
candy["Almond Joy", "winpercent"]
```
#Almond Joy; winpercent value: 50.35

>Q4. What is the winpercent value for "Kit Kat"?

```{r}
candy["Kit Kat", "winpercent"]
```
#winpercent value: 76.77

>Q5. What is the winpercent value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```
#Winpercent value: 49.65

```{r}
#install.packages("skimr)
library("skimr")
skim_without_charts(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

#The last column, winpercent, does not have a number in between 0 and 1.

>Q7. What do you think a zero and one represent for the candy$chocolate column?

#Zero is the candy does not contain chocolate and one is that it does contain chocolate.

>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy, aes(x=winpercent))+geom_histogram()
```

>Q9. Is the distribution of winpercent values symmetrical?

#Distribution is not symmetrical, seems to be skewed right

>Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
median(candy$winpercent)
```
#Center is below 50% of the distribution. 

>Q11. On average, is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate<-candy$winpercent[as.logical(candy$chocolate)]
fruity<-candy$winpercent[as.logical(candy$fruity)]
mean(chocolate)
mean(fruity)
```
#Chocolate is ranked higher than fruit candy with a mean average of 60.9% which is higher than the fruity candy (mean average of 44.1%).

>Q12. Is this difference statistically significant?

```{r}
t.test(chocolate, fruity)
```

#Yes, it is statistically significant. P-value is 2.871e-08 which is smaller than 0.05.

## Overall Candy Rankings

>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```
#Five least liked candy types are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbreakers.

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing=TRUE),], n=5)
```
#Top 5 all time favorite candy types in the set are Reese's Peanut Butter Cups, Reese's miniatures, Twix, Kit Kat, and Snickers. 

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col() + theme(axis.text.y=element_text(size=4))
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) + geom_col() + theme(axis.text.y=element_text(size=4))
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) + theme(axis.text.y=element_text(size=4))
```

>Q17. What is the worst ranked chocolate candy?

#Worst ranked chocolate candy is Sixlets.

>Q18. What is the best ranked fruity candy?

#Best ranked fruity candy is Starburst.

## Taking a look at pricepercent

```{r}
library(ggrepel)
ggplot(candy) + aes(winpercent, pricepercent, label=rownames(candy)) + geom_point(col=my_cols) + geom_text_repel(col=my_cols, size=3.3,max.overlaps=5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money -i.e. offers the most bang for your buck?

```{r}
cheap_candy<-candy[candy$pricepercent<0.3,]
cheap_candy[which.max(cheap_candy$winpercent),]
```

#Reese's Miniatures 

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
expensive <-head(candy[order(candy$pricepercent, decreasing=TRUE),],n=5)
expensive

expensive[which.min(expensive$winpercent),]
```
#or

```{r}
ord<-order(candy$pricepercent, decreasing=TRUE)
head(candy[ord, c(11,12)],n=5)
```
#The top five most expensive candy types in the dataset are Nik L Nip, Nestle Smarties, Ring pop, Hersheys Krackel, and Hersheys Milk Chocolate. The least popular out of them all is Nik L Nip.

>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called "dot chat" or "lollipop" chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) + aes(pricepercent, reorder(rownames(candy), pricepercent))+ geom_segment(aes(yend=reorder(rownames(candy), pricepercent),xend=0), col="gray40") + geom_point()+ theme(axis.text.y=element_text(size=4))
```

## Exploring the correlation structure

```{r}
library(corrplot)
```

```{r}
cij<-cor(candy)
corrplot(cij) + theme(axis.text.y=element_text(size=6))
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

#Chocolate and fruity are anti-correlated.

>Q23. Similarly, what two variables are more positively correlated?

#Chocolate and winpercent , chocolate and bar are most positively correlated.

## Principal Component Analysis

```{r}
pca<-prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2],col=my_cols,pch=16)
```

```{r}
my_data<-cbind(candy,pca$x[,1:3])
```

```{r}
p<-ggplot(my_data)+
  aes(x=PC1,y=PC2, size=winpercent/100,text=rownames(my_data), label=rownames(my_data)) + geom_point(col=my_cols) + theme(axis.text.y=element_text(size=6))
p
```

```{r}
library(ggrepel)

p+geom_text_repel(size=3.3, col=my_cols, max.overlaps=7) + theme(legend.position="none") + labs(title="Halloween Candy PCA Space", subtitle="Colored by type: chocolate bar(dark brown),chocolate otehr(light brown), fruity(red), other(black", caption="Data from 538") + theme(axis.text.y=element_text(size=6))
```

```{r}
#install.packages("plotly)
library(plotly)
```

```{r}
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2,ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

#Fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. This makes sense as this captures the spectrum between fruity and chocolate. The high PC1 is the fruity, hard candies the come in bulk, while the low PC1 is chocolate bars that are one item and is more expensive.


